Solving the Shallow Water Equations\\ using Finite Volumes and Lax-Friedrichs

### 2017-12-23 paper/paper.tex
Extracted code to appendix which uses minted

\input{paper-layout}
\input{german}
\usepackage{amsmath}
\pagestyle{empty}  % Skip page number on first page

@@

\begin{itemize}
\item In the 1-D case, the flow is in a channel of unit width
\item Horizontal velocity $u(x)$ across any cross section $x$ is constant
\item Vertical velocity $v(x)$ is small (but not zero) and falls out of the equations
\item Pressure is dominated by hydrostatics: $p=\frac{1}{2}\rho gh^2$
\end{itemize}

@@

This paper chooses the state variables $q := (h, hu)$ to form a state space, $Q$. This can be expressed in code as follows:
\begin{verbatim}
const Float = Float64

mutable struct Q
    h :: Float
    hu :: Float
end
\end{verbatim}

@@

This latter problem may be resolved quickly, thanks to Julia's type system.\cite{doi:10.1137/141000671}

\begin{verbatim}
import Base.+
import Base.-
import Base.*

function +(lhs::Q, rhs::Q)
    Q(lhs.h + rhs.h, lhs.hu + rhs.hu)
end

function -(lhs::Q, rhs::Q)
    Q(lhs.h - rhs.h, lhs.hu - rhs.hu)
end

function *(a::Real, q::Q)
    Q(a*q.h, a*q.hu)
end
\end{verbatim}

@@

This flux function may be expressed in Julia directly:

\begin{verbatim}
const G = 9.81f0
function f(q :: Q)
    Q(q.hu, q.hu^2/q.h + 0.5*G*q.h^2)
end
\end{verbatim}

The shallow water equations enter the numerical code in only one other place, the max wave speed function. The physical speeds of wave propagation are given by the eigenvalues of the flux Jacobian,

@@

will be shown to play an important role in ensuring numerical stability. In Julia it is calculated as:

\begin{verbatim}
function wavespeed(q :: Q)
    u = q.hu / q.h
    c = sqrt(G * q.h)
    max(abs(u-c), abs(u+c))
end
\end{verbatim}

@@

Let us now consider a clean organizing principle for the code.

Thanks to parametric polymorphism, the spatial domain can be quickly defined as a type alias \verb!QS = Array{Q,1}! This states that spatial variables are implemented as a 1-dimensional array with inner type \verb!Q! as defined above. There is an additional constraint in that there is a ghost cell on either side, so the `true' spatial domain ranges over the indices $[2, n_{cells}+1]$.

@@

In Julia, the CFL condition is encoded as follows. One useful feature is the function chain operator \verb!|>!, which aids in readability.

\begin{verbatim}
function cfl_dt(qs::QS, dx::Float, mu::Float)
    lambda = map(wavespeed,qs) |> maximum
    dx / lambda * mu
end
\end{verbatim}

\section{Model Problems}

@@

This can be encoded as a function 
* Input:  Buffers for $q(x), F_l(x), F_r(x)$, cell and timestep sizes
* Output: Updated inter-cell flux buffers

\begin{verbatim}
function lxf!(qs::QS, Fl::QS, Fr::QS, 
			  ncells::Int, 
              dx::Float, dt::Float)

    a = dx/dt
    for x = 2:ncells+2
        ql,qr = qs[x-1], qs[x]
        fl,fr = f(ql), f(qr)

        Fr[x-1] = 0.5*((fr-fl) - a*(qr - ql))
        Fl[x-1] = 0.5*((fr-fl) + a*(qr - ql))
    end
end
\end{verbatim}

@@

The experiment depicted in Figure~\ref{fig:lxf_gaussian_ncells} is rather intuitive. A Gaussian perturbation propagates symmetrically across the spatial domain. As the spatial discretization is refined, a shock wave gradually emerges from each (previously symmetric) ripple. It should be noted that these shock waves are smeared across tens or hundreds of cells, far more than the finite volume method presumably requires. Furthermore, the crest is reduced and the troughs are filled in: Decreasing the spatial discretization appears to increase dissipation.

@@

$a_{i-1/2} = \max(\lvert f'(q) \rvert) \ \forall q \in (Q_{i-1}, Q_{i})$

Then $ F_{i-1/2} = \frac{1}{2}[f(Q_{i-1}) + f(Q_i) - a_{i-1/2}(Q_i - Q_{i-1})]$

@@

This is represented in code as follows. The flux

\begin{verbatim}
function llxf!(qs::QS, Fl::QS, Fr::QS, 
              ncells::Int, 
              dx::Float, dt::Float)

    for x = 2:ncells+2
        ql,qr = qs[x-1], qs[x]
        fl,fr = f(ql), f(qr)

        a = max(wavespeed(ql), wavespeed(qr))

        Fr[x-1] = 0.5*((fr-fl) - a*(qr-ql))
        Fl[x-1] = 0.5*((fr-fl) + a*(qr-ql))
    end
end
\end{verbatim}

@@

%\input{conclusion}

@@

% can use a bibliography generated by BibTeX as a .bbl file
% standard IEEE bibliography style from:
% http://www.ctan.org/tex-archive/macros/latex/contrib/supported/IEEEtran/bibtex
\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
\bibliography{IEEEabrv,../references}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
%\begin{thebibliography}{1}
%
%\bibitem{ref:kopka}
%H.~Kopka and P.~W. Daly, \emph{A Guide to {\LaTeX}}, 3rd~ed.\hskip 1em plus
%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
%
%\end{thebibliography}

>>>>

\input{paper-layout}
\usepackage[outputdir=build]{minted}
\usepackage{tcolorbox}
\tcbuselibrary{minted,skins}

\newtcblisting{my_julia}[1][]{%
  listing engine=minted,
  minted language=julia,
  minted options={breaklines,breakanywhere,fontsize=\scriptsize,gobble=8},
  listing only,
  before skip=0pt,
  after skip=8pt,
  left skip=0pt,
  right skip=0pt,
  size=fbox,
  sharp corners,
  %colframe=white!75!black,
  colframe=white,
  boxrule=0pt,
  frame hidden,
  #1
}

\input{german}

\usepackage{amsmath}

\pagestyle{empty}  % Skip page number on first page

@@

\begin{itemize}
\item In the 1-D case, the flow is in a channel of unit width.
\item Horizontal velocity $u(x)$ across any cross section $x$ is constant.
\item Vertical velocity $v(x)$ is small (but not zero) and falls out of the equations.
\item Pressure is dominated by hydrostatics: $p=\frac{1}{2}\rho gh^2$
\end{itemize}

@@

This paper chooses the state variables $q := (h, hu)$ to form a state space, $Q$. As shown in lines XX-XX, this can be expressed in code as a lightweight mutable structure. The choice of single versus double precision for calculations can be specified via a single type alias.

@@

This latter problem may be resolved quickly, thanks to Julia's type system.\cite{doi:10.1137/141000671} Rather than having full objects, Julia provides container-like types. Methods have multiple dispatch, which means that a function is resolved to a method by examining all of the types in the signature, rather than simply the `owning' object. This means that operations such addition can be extended to custom types in a streamlined way. This can be seen in lines XX-XX.

Using parametric polymorphism, the spatial domain can be quickly defined as a type alias \verb!QS = Array{Q,1}! This states that spatial variables are implemented as a 1-dimensional array with inner type \verb!Q! as defined above. There is an additional constraint in that there is a ghost cell on either side, so the `true' spatial domain ranges over the indices $[2, n_{cells}+1]$.

@@

This flux function is expressed in lines XX-XX. The shallow water equations enter the numerical code in only one other place, the max wave speed function. The physical speeds of wave propagation are given by the eigenvalues of the flux Jacobian,

@@

will be shown to play an important role in ensuring numerical stability. This is implemented on lines XX-XX.

@@

Let us now consider a clean organizing principle for the code. A basic skeleton for running experiments is given on lines XX-XX. The following components are allowed to vary:

@@

The CFL condition is encoded on lines XX-XX. One useful feature is the function chain operator \verb!|>!, which aids in readability.

@@

@@

The experiment depicted in Figure~\ref{fig:lxf_gaussian_ncells} is rather intuitive. The Gaussian perturbation propagates symmetrically across the spatial domain. As the spatial discretization is refined, a shock wave gradually emerges from each (previously symmetric) ripple. It should be noted that these shock waves are smeared across tens or hundreds of cells, far more than the finite volume method presumably requires. Furthermore, the crest is reduced and the troughs are filled in: Decreasing the spatial discretization appears to increase dissipation.

@@

The viscosity parameter $a$ is taken to be the maximum of the wavespeeds of each neighboring cell. As shown in Figure~\ref{fig:artificial_viscosity}, this exhibits the desired behavior of being large in regions which are discontinuous or evolving quickly, and being small elsewhere.

\[a_{i-1/2} = \max(\lvert f'(q) \rvert) \ \forall q \in (Q_{i-1}, Q_{i})\]

\[ F_{i-1/2} = \frac{1}{2}[f(Q_{i-1}) + f(Q_i) - a_{i-1/2}(Q_i - Q_{i-1})] \]

@@

@@

@@

% can use a bibliography generated by BibTeX as a .bbl file
% standard IEEE bibliography style from:
% http://www.ctan.org/tex-archive/macros/latex/contrib/supported/IEEEtran/bibtex
\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
\bibliography{IEEEabrv,../references}

\input{appendix}

%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
%\begin{thebibliography}{1}
%
%\bibitem{ref:kopka}
%H.~Kopka and P.~W. Daly, \emph{A Guide to {\LaTeX}}, 3rd~ed.\hskip 1em plus
%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
%
%\end{thebibliography}
